---

title: "QDA Sensory Analysis with SensoryToolsR"
author: "Mai Nga"
date: "`r Sys.Date()`"
output:
rmarkdown::html_vignette:
toc: true
number_sections: true
vignette: >
%\VignetteIndexEntry{QDA Sensory Analysis with SensoryToolsR}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
-------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)

library(SensoryToolsR)
```

# Introduction

Quantitative Descriptive Analysis (QDA) is widely used to characterize and compare products using a trained sensory panel.

A typical replicated QDA experiment may include:

* several products;
* multiple trained assessors;
* repeated sessions or replicates;
* several sensory attributes measured on numerical intensity scales.

Analysis of such data commonly involves several complementary questions:

1. Is the experimental structure complete?
2. Do products differ in sensory intensity?
3. Do assessors discriminate among products?
4. Are assessors repeatable across sessions?
5. Is there evidence of Product × Assessor interaction?
6. Which sensory attributes contribute most strongly to product differentiation?
7. Can the multidimensional sensory structure be summarized using principal component analysis (PCA)?

SensoryToolsR provides functions that support these stages within a reproducible R workflow.

This vignette demonstrates the workflow using the included `qda_example` dataset.

# Example dataset

The package includes a simulated replicated QDA dataset:

```{r load-data}
data(qda_example)

dim(qda_example)
```

The dataset contains 108 observations:

* 6 products;
* 6 assessors;
* 3 sessions;
* 8 sensory attributes.

The first observations can be inspected using:

```{r inspect-data}
head(qda_example)
```

The experimental design variables are:

* `product`;
* `assessor`;
* `session`.

The sensory attributes are:

```{r attributes}
sensory_attributes <- c(
  "sweetness",
  "bitterness",
  "umami",
  "fishy_odor",
  "fresh_odor",
  "aftertaste",
  "firmness",
  "juiciness"
)

sensory_attributes
```

The dataset is simulated specifically for demonstration, teaching, examples, and package testing. It does not represent measurements obtained from human participants or a particular commercial product.

# Validate the sensory dataset

Before performing statistical analysis, the experimental structure should be checked.

SensoryToolsR provides `sensory_validate()` for this purpose.

```{r validation}
validation <- sensory_validate(
  qda_example
)
```

The validation procedure examines:

* number of rows and columns;
* number of assessors;
* number of products;
* number of sessions;
* sensory attributes;
* missing values;
* duplicate experimental records;
* expected assessor-product-session combinations;
* panel completeness.

For the example dataset, the expected number of design records is:

[
6\ \text{assessors}
\times
6\ \text{products}
\times
3\ \text{sessions}
==================

108
]

The dataset therefore represents a complete balanced assessor-product-session design.

# Descriptive statistics

Descriptive statistics provide an initial overview of the sensory profiles of the products.

Use `sensory_summary()` to calculate product-level summaries:

```{r sensory-summary}
sensory_summary_result <- sensory_summary(
  qda_example,
  attributes = sensory_attributes
)

sensory_summary_result
```

The output contains product-by-attribute descriptive information such as:

* number of observations;
* mean;
* standard deviation;
* standard error;
* median;
* coefficient of variation;
* confidence intervals;
* minimum;
* maximum.

These summaries are useful for understanding the magnitude and variability of sensory scores before performing inferential analyses.

# Analysis of one sensory attribute

A single sensory attribute can be analysed using `sensory_anova()`.

For example:

```{r single-anova}
sweetness_anova <- sensory_anova(
  qda_example,
  attribute = "sweetness"
)

sweetness_anova$anova_table
```

This analysis evaluates product differences while accounting for assessor effects.

The product p-value can be accessed directly:

```{r product-p-value}
sweetness_anova$product_p_value
```

A small product p-value provides evidence that the products differ in perceived sweetness.

# Post-hoc product comparisons

When the product effect is significant, pairwise product comparisons can be investigated using `sensory_posthoc()`.

```{r posthoc}
sweetness_posthoc <- sensory_posthoc(
  sweetness_anova
)

sweetness_posthoc$comparisons
```

SensoryToolsR currently uses Tukey HSD for these multiple comparisons.

Post-hoc analysis should normally be interpreted after considering the corresponding ANOVA result.

# Replicated panel ANOVA

Replicated sensory studies contain additional information about assessor and session performance.

Use `sensory_panel_anova()`:

```{r panel-anova}
sweetness_panel <- sensory_panel_anova(
  qda_example,
  attribute = "sweetness"
)

sweetness_panel$anova_table
```

The panel ANOVA evaluates:

* Product;
* Assessor;
* Session;
* Product × Assessor interaction.

## Product effect

A significant product effect indicates that the panel differentiates among products for the sensory attribute.

## Assessor effect

An assessor effect indicates differences in overall scale use among assessors.

Such an effect does not necessarily imply poor panel performance. Assessors may use different parts of the scale while still ranking products consistently.

## Session effect

A session effect can indicate systematic differences among repeated evaluation sessions.

## Product × Assessor interaction

The Product × Assessor interaction is particularly important in trained-panel evaluation.

A significant interaction may indicate that assessors do not perceive or rank product differences consistently.

The interaction therefore provides information about panel agreement in product discrimination.

# Individual assessor performance

SensoryToolsR can evaluate assessor-level performance using `sensory_panel_performance()`.

```{r assessor-performance}
sweetness_performance <- sensory_panel_performance(
  qda_example,
  attribute = "sweetness"
)

sweetness_performance$assessor_summary
```

The performance analysis considers multiple aspects of assessor behaviour rather than relying on a single statistic.

Depending on the dataset and selected thresholds, assessors may be identified for additional review.

A review flag should not automatically be interpreted as a reason to remove an assessor. Instead, it indicates that the assessor's data deserve closer inspection.

Possible considerations include:

* discrimination;
* repeatability;
* agreement with the panel;
* unusual scale use;
* inconsistent responses across sessions.

# Multi-attribute panel analysis

QDA studies normally contain several sensory attributes.

Instead of analysing every attribute manually, `sensory_panel_multi()` applies the panel analysis across multiple variables:

```{r panel-multi}
panel_multi <- sensory_panel_multi(
  qda_example,
  attributes = sensory_attributes
)

panel_multi$attribute_summary
```

This provides an attribute-level overview of:

* product significance;
* assessor effects;
* session effects;
* Product × Assessor interactions;
* assessor-performance screening.

The multi-attribute result is particularly useful for identifying which sensory attributes show strong product discrimination and which may require additional panel review.

# Visualizing panel performance

Panel-performance results can also be visualized.

For example:

```{r panel-plot, fig.alt="Sensory panel performance plot for sweetness"}
sensory_panel_plot(
  sweetness_performance
)
```

The graphical output complements the numerical panel-performance summaries and can help identify assessor patterns requiring further investigation.

# Principal component analysis

QDA data are multivariate because several sensory attributes jointly describe each product.

Principal component analysis provides a lower-dimensional representation of these sensory profiles.

Use:

```{r pca}
pca_result <- sensory_pca(
  qda_example,
  attributes = sensory_attributes,
  scale = FALSE
)

pca_result$variance_table
```

For `qda_example`, the first two principal components capture almost all of the systematic product-profile variation.

The simulated data were deliberately designed to contain two interpretable dimensions.

## PC1: taste and odor differentiation

PC1 is primarily associated with attributes related to freshness and deterioration.

Attributes such as:

* sweetness;
* fresh odor;
* umami;

tend to contrast with:

* bitterness;
* fishy odor;
* aftertaste.

The signs of PCA loadings may reverse without changing the scientific interpretation. Therefore, interpretation should focus on relative directions and magnitudes rather than on whether a loading is specifically positive or negative.

## PC2: texture differentiation

PC2 is primarily associated with:

* firmness;
* juiciness.

These two texture attributes occur in opposite directions on the second component.

This gives PC2 an interpretation that is largely independent of the freshness-related dimension represented by PC1.

# PCA biplot

A PCA biplot displays products and sensory attributes in the same two-dimensional representation.

```{r pca-biplot, fig.alt="PCA biplot of the QDA example dataset"}
sensory_pca_plot(
  pca_result,
  type = "biplot"
)
```

Products positioned close together have relatively similar sensory profiles.

Sensory attribute vectors indicate the directions associated with increasing values of those attributes.

Attributes pointing in similar directions tend to be positively associated, whereas attributes pointing in approximately opposite directions tend to represent contrasting product characteristics.

# PCA diagnostics

The statistical meaning of the PCA can be examined more systematically using `sensory_pca_diagnostics()`.

```{r pca-diagnostics}
pca_diagnostics <- sensory_pca_diagnostics(
  pca_result,
  components = c(1, 2),
  top_n = 5
)

pca_diagnostics$top_attributes
```

The top attributes table identifies the variables contributing most strongly to each principal component.

Product contributions can also be inspected:

```{r pca-products}
pca_diagnostics$top_products
```

Additional attribute-level diagnostics are available through:

```{r pca-attribute-diagnostics}
pca_diagnostics$attribute_diagnostics
```

These quantities help distinguish between:

* attributes strongly contributing to a component;
* attributes well represented by a component;
* products strongly influencing the PCA solution.

# Standardized versus non-standardized PCA

SensoryToolsR allows PCA either with or without standardization.

The previous analysis used:

```r
scale = FALSE
```

A standardized analysis can be performed using:

```{r standardized-pca}
pca_scaled <- sensory_pca(
  qda_example,
  attributes = sensory_attributes,
  scale = TRUE
)

pca_scaled$variance_table
```

Standardization gives each sensory attribute unit variance before PCA.

Whether standardization is appropriate depends on the scientific context.

When sensory attributes are measured on the same intensity scale and have comparable variability, non-standardized PCA may preserve meaningful differences in variance among attributes.

When attributes have substantially different scales or variances, standardized PCA may be more appropriate.

The choice should therefore be made explicitly rather than automatically.

# Integrated QDA analysis

The previous sections demonstrated individual SensoryToolsR functions.

For routine QDA analysis, these steps can be combined using `sensory_qda()`.

```{r integrated-qda}
qda_result <- sensory_qda(
  qda_example,
  attributes = sensory_attributes
)
```

Printing the result provides a concise integrated overview:

```{r print-qda}
qda_result
```

The result contains several components:

```{r qda-components}
names(qda_result)
```

Important components include:

```r
qda_result$overview
qda_result$validation
qda_result$summary
qda_result$panel
qda_result$pca
qda_result$pca_diagnostics
```

The integrated function does not introduce a separate statistical model.

Instead, it coordinates the validated SensoryToolsR analysis functions into a reproducible QDA workflow.

# Structured QDA summary

A structured summary can be created with:

```{r qda-summary}
qda_summary <- summary(
  qda_result
)

qda_summary$attribute_summary
```

The assessor-screening overview is available through:

```{r assessor-summary}
qda_summary$assessor_summary
```

The PCA summary can be inspected using:

```{r qda-pca-summary}
qda_summary$pca_variance
```

and the strongest PCA attribute contributors using:

```{r qda-top-attributes}
qda_summary$pca_top_attributes
```

This structured output is useful for downstream reporting because the statistical results remain available as tables rather than being limited to console text.

# A complete reproducible workflow

A compact QDA workflow using SensoryToolsR can therefore be written as:

```{r complete-workflow, eval=FALSE}
library(SensoryToolsR)

data(qda_example)

attributes <- c(
  "sweetness",
  "bitterness",
  "umami",
  "fishy_odor",
  "fresh_odor",
  "aftertaste",
  "firmness",
  "juiciness"
)

# Validate experimental structure
sensory_validate(
  qda_example
)

# Descriptive statistics
sensory_summary(
  qda_example,
  attributes = attributes
)

# Integrated QDA analysis
result <- sensory_qda(
  qda_example,
  attributes = attributes
)

# Concise result
result

# Structured summary
result_summary <- summary(
  result
)

# PCA visualization
sensory_pca_plot(
  result$pca,
  type = "biplot"
)

# PCA drivers
result$pca_diagnostics$top_attributes
```

# Interpretation principles

Statistical output should always be interpreted together with sensory-science knowledge and the experimental design.

Some useful principles are:

1. **Product significance indicates discrimination, not necessarily practical importance.**
   Statistical significance should be considered together with the magnitude of sensory differences.

2. **Assessor effects are not automatically problematic.**
   Assessors may use different portions of an intensity scale while maintaining similar product discrimination.

3. **Product × Assessor interaction deserves careful attention.**
   Strong interaction may indicate disagreement in product characterization.

4. **Assessor review flags are diagnostic, not automatic exclusion rules.**
   Decisions about assessor retraining or exclusion require examination of the original study context.

5. **PCA axes should be interpreted from loadings and contributions.**
   Component signs are arbitrary and may reverse without changing the sensory meaning.

6. **PCA scaling should be a scientific decision.**
   Standardization should not be applied automatically.

7. **Statistical analysis does not replace good sensory experimental design.**
   Appropriate panel selection, training, attribute development, replication, sample presentation, and testing conditions remain essential.

# Reproducibility

A reproducible sensory workflow should retain:

* raw sensory data;
* cleaned data;
* experimental design information;
* analysis scripts;
* package versions;
* analysis settings;
* statistical outputs;
* graphical outputs.

The SensoryToolsR workflow is designed to help make these steps explicit and repeatable.

For formal research, it is also good practice to record the R session information:

```{r session-info}
sessionInfo()
```

# Summary

SensoryToolsR provides an integrated framework for analysing replicated QDA sensory data.

The workflow demonstrated in this vignette includes:

1. experimental-structure validation;
2. descriptive sensory statistics;
3. single-attribute ANOVA;
4. post-hoc product comparisons;
5. replicated panel ANOVA;
6. assessor-performance assessment;
7. multi-attribute panel analysis;
8. panel visualization;
9. principal component analysis;
10. PCA diagnostics;
11. integrated QDA analysis;
12. structured QDA summaries.

The included `qda_example` dataset allows the complete workflow to be reproduced directly after installing the package.
